Seminar: Programming models and Code Generation\\ Winter Term 2014/2015\\ {\bf Quasi deterministic parallel programming using LVars}

### 2015-01-09 draft.tex
new

To keep track of the running callbacks, LVish includes an auxiliary form of the above pattern taking additional arguments: the set of running callbacks $B$ and a set of values $H$ in the event set for which callbacks have already been launched. It detects quiescence by checking: Every event of interest that has occurred must be handled (be in $H$). Second, all existing callback threads must have completed and terminated with a value. If yes, then LVar state is freezed. To ensure quiescence is a permanent state, rather than a transient one, freezing is usually the very last step of an algorithm, permitting its result to be extracted. The $runParThenFreeze$ function inside LVish library does this, and guarantees full determinism.\\ \\
\textbf{\textit{Proof of Quasi determinism for Lvish:}} According to \cite{lkuper2} the main theorem says that if two executions starting from a configuration $\sigma$ terminate in configurations $\sigma{'}$ and $\sigma{''}$, then either $\sigma{'}$ and $\sigma{''}$ are the same configuration (up to a permutation on locations), or one of them is \emph{error}.\\
\textit{Independence property:} Like LVars, the LVish model uses the frame property again to capture the idea that independent effects commute with each other but this time with some restrictions. If we know that a \emph{store} $S$ evaluating expression $e$ steps to a new \emph{store} $S{'}$ with $e{'}$ then it is valid to say that $S\sqcup_{S}S''$ evaluating $e$ steps to a new store $S'\sqcup_{S}S''$ with $e'$ if the following conditions are satisfied:
\begin{itemize}
\item The status bits of $S'\sqcup_{S}S''$ and S agree with each other for all locations shared between them.
\item Locations in $S''$ cannot share the same name with locations newly allocated during the transition from one configuration
to another.
\end{itemize}
For details on lemmas and their proofs, refer \cite{kuper}\cite{lkuper2}.

@@

It adopts the basic approach of the $par$ monad. The existing $par$ monad\cite{simon} in Haskell offers an API for parallel programming. It providess by default a work-stealing scheduler and supports forking tasks. The programmer specifies how information flows across computations but not the order in which they will be evaluated at runtime. Information flow is described using IVars. With LVish we get a lattice-generic infrastructure: the $par$ monad, a thread scheduler, support for blocking and signaling threads, handler pools, and event handlers. This infrastructure does not guarantee quasi-determinism, only data structure authors should import it who will implement a specific monotonic data structure using this library providing efficient parallel access, exporting a limited interface for its users ensuring quasi-determinism. LVish library provides its own custom scheduler and programs run only inside the $par$ monad allowing only Lvish-sanctioned side effects to provide compile time guarantees of determinism and quasi determinism.\\ \\

@@

\textbf{\textit{Key Ideas:}} \textit{Atoms and Idempotence:} Atoms are elements not equal to $\bot$ but whose only smaller
element is $\bot$. Lattices in which every element is the LUB of some set of atoms are called atomistic, and most application-specific lattices used by LVish programs have this property. \ref{Figure1} is an example for an atomistic lattice where the atoms are singleton elements and all the elements can be obtained as LUB of these atoms. Data structures corresponding to such type of lattices usually expose operations that work at the atom level, semantically limiting $puts$ to atoms, $gets$ to threshold sets of atoms, and event sets to sets of atoms. LVish facilitates good performance for atomistic lattices by associating LVars with a set of deltas(changes), as well a lattice. For atomistic lattices, the deltas are just the atoms, for a set, a delta is an element; for a map, a key/value pair. Deltas allow easy and efficient communication of changes between $puts$ and $gets$/handlers. \par
\textit{Idempotence,} meaning that $d\sqcup{d=d}$ for any element $d$. Repeated $puts$ or $freezes$ have no effect, if the scheduler is allowed to occasionally duplicate work, it is possible to save on synchronization costs. LVish computations are guaranteed to be idempotent and uses the scheduler(\textit{standard Chase-Lev deque}). Idempotence helps to deal with races between $put$ and $get/addHandler$. \par 

@@

\textbf{\textit{Limitations and Related work:}} The feasibility of LVar threshold reads have been explored in a distributed

@@

setting\cite{newton2}. Distributed systems involve replication of data objects across several locations making them robust to data loss and provide better data locality. Eventual consistency is a policy where the updates for data objects \textbf{\textit{eventually}} reach the replicas, that need not always agree on queries. Convergent replicated data types provide a mathematical framework for reasoning about and enforcing the eventual consistency of such replicated objects, based on viewing them as elements of a lattice and replica conflict resolution as the lattice's join operation. CvRDTs are provably eventually consistent, but queries of CvRDTs allow inconsistent intermediate states of replicas to be observed. LVar-style threshold queries to CvRDTs extend them with support for deterministic queries and show that they are strongly consistent queries. A threshold query that returns an answer when executed on a replica will return the same answer every subsequent time, and on a different replica, it will eventually return the same answer, otherwise will block until it does so. It is therefore impossible to observe different results from the same threshold query, whether at different times on the same replica, or on different replicas.\par
\cite{newton1}discusses the viability of Lvish for certain graph algorithms, its drawbacks and enhancements. Lvish side effects are only monotonic updates. If a data structure needs to be updated multiple times, it will have to be done out of place by allocating LVars frequently, resulting in additional memory allocation and inefficiency relative to the in-place versions. Blocking $get$ semantics could slow down the performance until the dependency is met. Instead of blocking, the tasks could be retried at a later time/rescheduled when there dependency has been satisfied and the paper proposes an abort/retry feature for the same. The overhead in going through the scheduler many times has also been highlighted for applications like Maximum independent set.

>>>>

To keep track of the running callbacks, LVish includes an auxiliary form of the above pattern taking additional arguments: the set of running callbacks $B$ and a set of values $H$ in the event set for which callbacks have already been launched. It detects quiescence by checking: Every event of interest that has occurred must be handled (be in $H$). Second, all existing callback threads must have completed and terminated with a value. If yes, then LVar state is freezed. To ensure quiescence is a permanent state, rather than a transient one, freezing is usually the very last step of an algorithm, permitting its result to be extracted. The $runParThenFreeze$ function inside LVish library does this, and guarantees full determinism.\\ \\
\textbf{\textit{Proof of Quasi determinism for Lvish:}} According to \cite{lkuper2} the main theorem says that if two executions starting from a configuration $\sigma$ terminate in configurations $\sigma{'}$ and $\sigma{''}$, then either $\sigma{'}$ and $\sigma{''}$ are the same configuration (up to a permutation on locations), or one of them is \emph{error}.\\
\textit{Independence property:} Like LVars, the LVish model uses the frame property again to capture the idea that independent effects commute with each other but this time with some restrictions. If we know that a \emph{store} $S$ evaluating expression $e$ steps to a new \emph{store} $S{'}$ with $e{'}$ then it is valid to say that $S\sqcup_{S}S''$ evaluating $e$ steps to a new store $S'\sqcup_{S}S''$ with $e'$ if the following conditions are satisfied:
\begin{itemize}
\item Status bits of $S'\sqcup_{S}S''$ and S agree with each other for all locations shared between them.
\item Locations in $S''$ cannot share the same name with locations newly allocated during the transition from one configuration
to another.
\end{itemize}
For details on lemmas and their proofs, refer \cite{kuper}\cite{lkuper2}.

@@

It adopts the basic approach of the $par$ monad. The existing $par$ monad\cite{simon} in Haskell offers an API for parallel programming. It providess by default a work-stealing scheduler and supports forking tasks. The programmer specifies how information flows across computations but not the order in which they will be evaluated at runtime. Information flow is described using IVars. With LVish we get a lattice-generic infrastructure: the $par$ monad, a thread scheduler, support for blocking and signaling threads, handler pools, and event handlers. This infrastructure does not guarantee quasi-determinism, only data structure authors should import it who will implement a specific monotonic data structure using this library providing efficient parallel access, exporting a limited interface for its users ensuring quasi-determinism. LVish library provides its own custom scheduler and programs run only inside the $par$ monad allowing only Lvish-sanctioned side effects to provide compile time guarantees of determinism and quasi determinism.\par

@@

\textbf{\textit{Key Ideas:}} \textit{Atoms and Idempotence:} \textit{Atoms} are elements not equal to $\bot$ but whose only smaller element is $\bot$. Lattices in which every element is the LUB of some set of atoms are called atomistic, and most application-specific lattices used by LVish programs have this property. \ref{Figure1} is an example for an atomistic lattice where the atoms are singleton elements and all the elements can be obtained as LUB of these atoms. Data structures corresponding to such type of lattices usually expose operations that work at the atom level, semantically limiting $puts$ to atoms, $gets$ to threshold sets of atoms, and event sets to sets of atoms. LVish facilitates good performance for atomistic lattices by associating LVars with a set of deltas(changes), as well a lattice. For atomistic lattices, the deltas are just the atoms, for a set, a delta is an element; for a map, a key/value pair. Deltas allow easy and efficient communication of changes between $puts$ and $gets$/handlers. \par
\textit{Idempotence,} meaning that $d\sqcup{d=d}$ for any element $d$. Repeated $puts$ or $freezes$ have no effect, if the scheduler is allowed to occasionally duplicate work, it is possible to save on synchronization costs. LVish computations are guaranteed to be idempotent and uses the scheduler(\textit{standard Chase-Lev deque}). Idempotence helps to deal with races between $put$ and $get/addHandler$. \\ \\ 

@@

\textbf{\textit{Limitations and Related work:}} \cite{newton1}discusses the viability of Lvish for certain graph algorithms, its

@@

drawbacks and enhancements. Lvish side effects are only monotonic updates. If a data structure needs to be updated multiple times, it will have to be done out of place by allocating LVars frequently, resulting in additional memory allocation and inefficiency relative to the in-place versions. Blocking $get$ semantics could slow down the performance until the dependency is met. Instead of blocking, the tasks could be retried at a later time/rescheduled when there dependency has been satisfied and the paper proposes an abort/retry feature for the same. The overhead in going through the scheduler many times has also been highlighted for applications like Maximum independent set.\\ \par
The feasibility of LVar threshold reads have been explored in a distributed setting\cite{newton2}. Distributed systems involve replication of data objects across several locations making them robust to data loss and provide better data locality. Eventual consistency is a policy where the updates for data objects \textbf{\textit{eventually}} reach the replicas, that need not always agree on queries. Convergent replicated data types provide a mathematical framework for reasoning about and enforcing the eventual consistency of such replicated objects, based on viewing them as elements of a lattice and replica conflict resolution as the lattice's join operation. CvRDTs are provably eventually consistent, but queries of CvRDTs allow inconsistent intermediate states of replicas to be observed. LVar-style threshold queries to CvRDTs extend them with support for deterministic queries and show that they are strongly consistent queries. A threshold query that returns an answer when executed on a replica will return the same answer every subsequent time, and on a different replica, it will eventually return the same answer, otherwise will block until it does so. It is therefore impossible to observe different results from the same threshold query, whether at different times on the same replica, or on different replicas.  
